---
title: "New CellPhoneDB v5 results"
output: rmarkdown::html_vignette
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{New CellPhoneDB v5 results}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/licenses/MIT)
[![codecov](https://codecov.io/gh/zktuong/ktplots/branch/master/graph/badge.svg)](https://codecov.io/gh/zktuong/ktplots)
[![R](https://github.com/zktuong/ktplots/actions/workflows/r.yml/badge.svg)](https://github.com/zktuong/ktplots/actions/workflows/r.yml)
[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6728350.svg)](https://doi.org/10.5281/zenodo.5717922)

# CellPhoneDB v5 results

From version 5 of [CellPhoneDB](https://www.github.com/ventolab/cellphonedb), there is a new output file - `interaction_scores`. 

According to the official repository, this table corresponds to:

`interaction_scores`: stores the new score generated. This score ranges from 0-100.
 
To score interactions CellPhoneDB v5 employs the following protocol:
 
1.    Exclude genes not participating in any interaction and those expressed in less than k% of cells within a given cell type.
2.    Calculate the mean expression (G) of each gene (i) within each cell type (j).
3.    For heteromeric proteins, aggregate the mean gene expression of each subunit (n) employing the geometric mean.
4.    Scale mean gene/heteromer expression across cell types between 0 and 100.
5.    Calculate the product of the scale mean expression of the interaction proteins as a proxy of the interaction relevance.

`cellsign`: accepts the new `CellSign` data. # TODO: ask Kevin for result description

`ktplots` will support these output via inclusion into the existing `plot_cpdb` function. We will gradually enable their functionality across the other functions, as well as with in the python package eventually.

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

### Load packages
```{r, message = FALSE, warning = FALSE}
library(ktplots)
data(cpdb_output_v5)

plot_cpdb_heatmap(pvals=relevant_interactions_v5, degs_analysis=TRUE, title="Sum of significant interactions")
```

```{r, message = FALSE, warning = FALSE, fig.height=3, fig.width=10}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11|PV STEAP4|PV MMPP11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    genes=c("TGFB2", "CSF1R"),
    title="Interactions between PV and trophoblast ",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE
)
```

### Interaction scores

Let's start with interaction scores. If a dataframe corresponding to the `interaction_scores` file is provided, you can toggle the alpha transparency of the interactions by the interaction score (interaction ranking is simply the score/100).

```{r, message = FALSE, warning = FALSE, fig.height=3, fig.width=12}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11|PV STEAP4|PV MMPP11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    genes=c("TGFB2", "CSF1R"),
    title="Interactions between PV and trophoblast ",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE,
    interaction_scores=interaction_scores_v5,
    scale_alpha_by_interaction_scores=TRUE
)
```

You can also specify a minimum interaction score to keep, removing all interactions lesser than this value.

```{r, message = FALSE, warning = FALSE, fig.height=3, fig.width=8}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11|PV STEAP4|PV MMPP11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    genes=c("TGFB2", "CSF1R"),
    title="Interactions between\nPV and trophoblast ",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE,
    interaction_scores=interaction_scores_v5,
    min_interaction_score=20
)
```

or specify both to have the alpha transparency shown too.

```{r, message = FALSE, warning = FALSE, fig.height=3, fig.width=10}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11|PV STEAP4|PV MMPP11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    genes=c("TGFB2", "CSF1R"),
    title="Interactions between\nPV and trophoblast ",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE,
    interaction_scores=interaction_scores_v5,
    scale_alpha_by_interaction_scores=TRUE,
    min_interaction_score=20
)
```

### CellSign

If a dataframe corresponding to the `cellsign` file is provided, you can toggle the filter the interactions by the results

```{r, message = FALSE, warning = FALSE, fig.height=4, fig.width=8}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    title="Interactions between\nPV and trophoblast with\ndownstream significance",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE,
    cellsign=cellsign_v5,
    filter_by_cellsign=TRUE
)
```

and also scale the alpha value (50% for 0 and 100% for 1).

```{r, message = FALSE, warning = FALSE, fig.height=4, fig.width=8}
plot_cpdb(
    scdata=sce_v5,
    cell_type1="PV MYH11",
    cell_type2="EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
    means=means_v5,
    pvals=relevant_interactions_v5,
    celltype_key="cell_labels",
    title="Interactions between\nPV and trophoblast with\ndownstream significance",
    max_size=6,
    highlight_size=0.75,
    degs_analysis=TRUE,
    standard_scale=TRUE,
    cellsign=cellsign_v5,
    filter_by_cellsign=TRUE,
    scale_alpha_by_cellsign=TRUE
)
```

### Additional plotting data

From now on, `is_integrin`, `directionality` and `classification` are transferred to final output table in `plot_cpdb`. This means you will be able to use something like `facet_grid`/`facet_wrap`` to plot them!

```{r, message = FALSE, warning = FALSE, fig.height=3, fig.width=20}
library(ggplot2)
p <- plot_cpdb(
        scdata = sce_v5,
        cell_type1 = "PV MYH11|PV STEAP4|PV MMPP11",
        cell_type2 = "EVT_1|EVT_2|GC|iEVT|eEVT|VCT_CCC",
        means = means_v5,
        pvals = relevant_interactions_v5,
        celltype_key = "cell_labels",
        genes = c("TGFB2", "CSF1R", "COL1A1"),
        max_size = 6,
        highlight_size = 0.75,
        degs_analysis = TRUE,
        standard_scale = TRUE,
        interaction_scores = interaction_scores_v5
        # return_table = TRUE
    )
p + facet_wrap(~classification, ncol = 3)
```
